Bifurcation or state tipping: assessing transition type in a model trophic cascade

### 2018-07-14 paper/paper.Rmd
add table, adj cite, cite batt2017 re:s-tipping

There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or bifurcation (B-tipping).  Figure 1 illustrates these two different mechanisms.  B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]). In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.  S-tipping can be driven by purely stochastic events or direct exogenous forcing. Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g. @Wissel1984; @Scheffer2009; @Scheffer2015].

@@

>>>>

There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or bifurcation (B-tipping).  Figure 1 illustrates these two different mechanisms.  B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]). In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.  S-tipping can be driven by purely stochastic events or direct exogenous forcing, such as extreme events that are common in the biological variables of ecosystems [@Batt2017]. Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g. @Wissel1984; @Scheffer2009; @Scheffer2015].

@@

**Table 1.** Model parameters, description, values, and units. Parameters are taken from [@carpenter2008] and [@Carpenter2013]. The function \\gamma describes phytoplankton growth, and can be found in [@carpenter2008].

|Symbol |Description                                                 |Quantity |Units               |
|:------|:-----------------------------------------------------------|:--------|:-------------------|
|q      |Harvest or stocking                                         |variable |t^{-1}              |
|s      |Survival rate of juveniles to maturation                    |0.6      |t^{-1}              |
|m_A    |Mortality rate of adults                                    |0.4      |t^{-1}              |
|m      |Phytoplankton mortality                                     |0.1      |t^{-1}              |
|f      |Fecundity of adult bass                                     |2        |J/A                 |
|c_{FA} |Predation of planktivores by adult bass                     |0.3      |t^{-1} A^{-1}       |
|c_{JA} |Predation of juvenile bass on by adult bass                 |0.1      |t^{-1} A^{-1}       |
|c_{JF} |Predation of juvenile bass by planktivorous fish            |0.5      |t^{-1} F^{-1}       |
|c_{PH} |Predation of phytoplankton by herbivores                    |0.25     |t^{-1} H^{-1}       |
|c_{HF} |Predation of herbivores by planktivores                     |0.1      |t^{-1} F^{-1}       |
|F_o    |Abundance of planktivorous fish in non-foraging arena       |200      |F                   |
|H_R    |Reservoir of herbivores in a deep-water refuge              |4        |H                   |
|D_F    |Diffusion of planktivores between refuge and foraging arena |0.09     |t^{-1}              |
|DH     |Diffusion of between shallow and deep water                 |0.5      |t^{-1}              |
|v      |Rate at which J enter a foraging arena, becoming vulnerable |80       |t^{-1}              |
|h      |Rate at which J hide in a refuge                            |80       |t^{-1}              |
|r_P    |Growth rate of phytoplankton                                |3        |m^2 mg^{-1}         |
|L      |Phosphorus loading rate                                     |0.6      |mg m^{-2} t^{-1}    |
|\alpha |Assimilation of phytoplankton phosphorus by zooplankton     |0.3      |g g^{-1}            |
|\gamma |Density-dependent phytoplankton growth response to light    |n/a      |n/a                 |
|I0     |Solar irradiance incident on the surface of the water       |300      |\mu E m^{-2} s^{-1} |
